T-reX: Quantifying waste and material footprints in current and future Life Cycle Assessment (LCA) databases

### 2024-11-30 00_JCR-submission-folder_SECOND-SUBMISSION/00_T-reX_manuscript.tex
update

\usepackage{booktabs}
%%%Author macros
\def\tsc#1{\csdef{#1}{\textsc{\lowercase{#1}}\xspace}}
\tsc{SCM}
\tsc{EL}
\tsc{SC}
\tsc{CFB}
%%%

\usepackage{caption}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
\usepackage[british]{babel}
\usepackage{microtype}
\usepackage{listings}
\usepackage{xcolor}
\definecolor{blue}{RGB}{33,150,209}
\definecolor{red}{RGB}{255,0,0}% \usepackage{listings}
\usepackage{xurl}
% \usepackage{pdfpages}
\usepackage{enumitem}
% \usepackage{lineno}
\usepackage{float}
% 
% \setlength{\linenumbersep}{5pt}
% Save the old \texttt command
\let\oldtexttt\texttt
% Redefine the \UrlFont to use a smaller typewriter font
\renewcommand{\UrlFont}{\ttfamily\small}
% Redefine the \texttt command to use a smaller font size and allow line breaks
\renewcommand{\texttt}[1]{{\ttfamily\small\nolinkurl{#1}}}
\usepackage{hyperref}

\usepackage{graphicx}

@@

	With a simple case study using Lithium-ion batteries, we demonstrate some of
	T-reX's potential by detailing and exploring aspects of their waste and
	material footprints. Since such footprints are often linked to negative
	externalities, T-reX could support sustainable decision-making and contribute
	to the development of the circular economy by promoting the understanding of
	our material consumption and waste generation.

\verb+\begin{keyword}+ \verb+\sep+circular economy \verb+\sep+waste \verb+\sep+material footprint \verb+\sep+life cycle assessment \verb+\sep+critical raw materials \verb+sep+supply chain \verb+\end{keyword}+

\end{abstract}

@@

@@

\maketitle

@@

% Introduction
\section{Introduction}\label{sec:introduction}
\subsection{Introduction to T-reX}\label{sec:intro-trex}

The development of a circular economy has become a central focus in the pursuit
of sustainability objectives toward curtailment of our environmental footprint
within planetary boundaries~\citep{eu2019greendeal,
	eu2020circ,nl2023ceplan,nl2016ceplan,pardo2018ce,ellenmacarthur2015ce}.
Fundamental to this development is a decrease in primary material consumption
and a reduction of life cycle waste through the implementation of `re-X'
strategies (e.g., refuse, rethink, design for---and implementation of---repair,
remanufacturing and recycling)~\citep{reike2018rex, eu2022ecodesign,
	eu2022repair, eu2015reman}. In addition to circular economy goals, contemporary
geopolitical tensions in an ever more globalised world have highlighted the
vulnerability of many advanced economies to intentional supply
disruptions---wrought as an act of competition or even outright
hostility~\citep{jrc2023supplychain,hartley2024cepolitics,berry2023crm}.

@@

Several LCIA methods include, to some extent, waste generation (Swiss
Eco-Factors, EDIP and
EN15804)~\citep{foen2021ecofactors,hauschild2003edip,cen2019en15804} and
material consumption (Crustal Scarcity Indicator (CSI) and Swiss
Eco-Factors~\citep{arvidsson2020csi,foen2021ecofactors}). These methods,
however, are generally limited in their scope (especially for waste), do not
allow for flexible quantification of specific waste and material types, and
often provide results in characterised units that are abstract or difficult to
interpret (e.g., \textit{Ümweltbelastungspunkte} (UBP) [English: environmental
	impact points] in the case of the Swiss
Eco-Factors)~\citep{su2020sustainableproddev}.

@@

% Methodology

@@

In the default configuration of T-reX, there are 10 waste categories which are
then further divided by their unit of measurement (kilograms and cubic meters)
to create a total of 20 waste methods. The waste categories include
aggregations such as 'total waste', 'hazardous waste' as well as, those for
waste sent to treatment by, for example, incineration or recycling. The full
details are listed in the supplementary material in~\autoref{sec:supplementary}
[section 3 ]. One advantage of T-reX is that it can identify `hidden' waste
exchanges that would otherwise be `consumed' by a treatment process and not
leave the technosphere.

@@

An overview of the T-reX workflow is presented in
\autoref{fig:methods-flowchart}. The supplementary material in
\autoref{sec:supplementary} (part 4) contains a more detailed computational flowchart.

\begin{figure*}
	\centering
	\caption{The workflow of T-reX presented in a generalised (a) and computational (b) format. Application of T-reX to one or more LCA databases creates a \texttt{brightway} project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}
	\includegraphics[width=1.8\columnwidth]{figures/T-reX_method.pdf}\label{fig:methods-flowchart}
\end{figure*}

@@

To provide an example, we investigated five types of Li-ion batteries in the
\texttt{ecoinvent} 3.9.1 `cutoff' LCA database~\citep{ecoinvent2016version3},
each represented by their unamended global market inventories. The system
boundaries. These following markets were used for the example calculations and
unaltered by the authors:
\begin{itemize}
	\itemsep=0pt
	\item Li-ion, LFP, rechargeable, prismatic
	\item Li-ion, LiMn\(_2\)O\(_4\), rechargeable, prismatic
	\item Li-ion, NMC111, rechargeable, prismatic
	\item Li-ion, NMC811, rechargeable, prismatic
	\item Li-ion, NCA, rechargeable, prismatic
\end{itemize}

@@

\begin{itemize}
	\itemsep=0pt
	\item ReCiPe 2016 v1.03, midpoint (I)~\citep{huijbregts2016recipe}
	\item Environmental Footprint (EF) v3.1 (no long-term emissions)~\citep{eu2023ef}
	\item Spatial differentiation in life cycle impact assessment---the EDIP 2003 no long
	      term emissions method~\citep{hauschild2003edip}
	\item The crustal scarcity indicator method---CSI 2020~\citep{arvidsson2020csi}
\end{itemize}

@@

An example of the output from the application of T-reX has been included in the
supplementary material in~\autoref{sec:supplementary}, appendix 6. The
manipulated \texttt{ecoinvent} databases (which are the main product of T-reX)
can be recreated by following the instructions in the package
documentation~\citep{mcdowall2023T-reXdocs}.

@@

\autoref{fig:sankey_waste} presents a heavily simplified Sankey diagram depicting the supply chain flows of `total solid waste' attributed to the NMC 811 Li-ion battery at the global market in the \texttt{ecoinvent} database version 3.9.1. As the supply chains are expansive, this visualisation has been greatly simplified to be legible for print, multiple versions are included in the supplementary material~\autoref{sec:supplementary}. As is evident in the figure, the largest contributors to the waste footprint are the extraction and processing of the raw materials, with the largest flows coming from the extraction of cobalt, lithium, and nickel. Due to the nature of LCA, the waste flows do not represent the `real' physical flows directly, but rather an abstraction thereof based on, in this case, economic allocation---the market price of the functional unit produced by the activity~\citep{guinee2004economicallocation}. If an unallocated database were available, the waste flows could be calculated with T-reX in the same way and would be more directly related to the physical flows of the activities.

\begin{figure}
	\centering
	\includegraphics[width=0.9\columnwidth]{figures/T-reX_NMC811_WasteTotalSolid.pdf}
	\caption{A greatly simplified Sankey diagram from \texttt{ActivityBrowser} showing the total solid waste footprint flows for the NMC 811 battery supply chain at the global market in the \texttt{ecoinvent} database version 3.9.1}\label{fig:sankey_waste}
\end{figure}

\subsubsection{Temporal and scenario variation in waste and material inventory footprints}
%\label{sec:results-case_study-total_footprints}

@@

\begin{figure*}[]
	\centering
	\includegraphics[width=2.1\columnwidth]{figures/T-reX-wastefootprint-totalANDcarbon.pdf}
	\caption{T-reX calculated supply chain inventories for the five Li-ion batteries in the case study, showing their `footprints' for: (a) total solid waste, and (b) carbon dioxide waste (produced from carbon capture and storage (CCS) only). The footprints were modelled with prospective LCA databases from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:waste_totalANDcarbon}
\end{figure*}
%!!!!

@@

\begin{figure*}
	\centering
	\includegraphics[width=2.0\columnwidth]{figures/T-reX-coalANDedip.pdf}
	\caption{Comparison of the (a) `T-reX  	\caption{Comparison of the (a) `T-reX  Coal (black)' demand footprint with the (b) conventional LCIA method `EDIP 2003 - coal no LT'  in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model for all five batteries.}\label{fig:comparison_methods}
Coal (black)' demand footprint with the (b) conventional LCIA method `EDIP 2003 - coal no LT'  in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model for all five batteries.}\label{fig:comparison_methods}
\end{figure*}

@@

In the most similar published method, \cite{laurenti2023wastefootprint} used an
alternative procedure for calculating the waste footprint of \~1400 activities
in the \texttt{ecoinvent} database version 3.5. Direct comparison is not
possible, as this database contains only one generic Li-ion battery, `market
for battery, Li-ion, rechargeable, prismatic'. The inventory of this battery
most closely resembles that of the NMC 111 battery in this case study.
\autoref{tab:results-case_study-comparison} presents a comparison of the
results from the two studies, where possible. For liquid, solid, and recycled
waste, the results were closely aligned, however, for the hazardous waste
fraction, Laurenti et al. reported 95\%, whereas T-reX reported only 3\%. The
reason for this discrepancy is explained by the fact that in the method of
Laurenti et al.\, a ``waste flow was regarded as hazardous when the
hazardousness was clearly stated in its subsequent waste treatment activity''.
The authors continue: ``It should be noted that the high hazardousness ratios
for many products might indicate a weakness in the validity of this measure''.
In T-reX, the source database is deconstructed into a list of separate
exchanges, and only those explicitly defined as hazardous are marked as such.

@@

% Back matter
%\appendix
\section{Supplementary material}\label{sec:supplementary}
The supplementary material supplied with this manuscript contains the
following:
\begin{enumerate}
	\item Metadata of the T-reX Python package
	\item Details of the modules and user configuration
	\item Description of the computational workflow
	\item Modular flowchart of T-reX
	\item Example of the terminal output of T-reX
	\item Example of the terminal output of the case study
	\item Complete tabulated results of the case study
	\item Complete visualisations from the case study
\end{enumerate}

@@

\includegraphics[width=0.9\columnwidth]{figures/T-reX_credit_heatmap.pdf}

+\verb+\printcredits +\verb+\credit+ \printcredits

@@

% insert supplementary information
%\clearpage
%\includepdf{T-reX_supplementary-material/00_T-reX_supplementary-material_main.pdf}

>>>>

%Author macros
\def\tsc#1{\csdef{#1}{\textsc{\lowercase{#1}}\xspace}}
\tsc{SCM}
\tsc{EL}
\tsc{SC}
\tsc{CFB}
%%%

% Essential packages
\usepackage{graphicx}
\usepackage{float} % For precise figure placement
\usepackage[authoryear]{natbib} % Use natbib for citations

% Optional packages (uncomment only if necessary)
%\usepackage{subcaption} % For subfigures
%\usepackage{xcolor} % If custom colours are needed
%\usepackage{lineno} % For line numbering (use \linenumbers to activate)

% Ensure compatibility
\usepackage{booktabs} % For professional-quality tables

% \setlength{\linenumbersep}{5pt}
%
\usepackage{graphicx}

@@

With a simple case study using Lithium-ion batteries, we demonstrate some of
T-reX's potential by detailing and exploring aspects of their waste and
material footprints. Since such footprints are often linked to negative
externalities, T-reX could support sustainable decision-making and contribute
to the development of the circular economy by promoting the understanding of
our material consumption and waste generation.
\end{abstract}

@@

\begin{keyword}
circular economy \sep wste \sep material footprint \sep life cycle assessment \sep critical raw materials \sep supply chain
\end{keyword}

@@

%\maketitle

@@

% Introduction
\section{Introduction}\label{sec:introduction}

\subsection{Introduction to T-reX}\label{sec:intro-trex}

The development of a circular economy has become a central focus in the pursuit
of sustainability objectives toward curtailment of our environmental footprint
within planetary boundaries~\citep{eu2019greendeal,	eu2020circ,nl2023ceplan,nl2016ceplan,pardo2018ce,ellenmacarthur2015ce}.

Fundamental to this development is a decrease in primary material consumption
and a reduction of life cycle waste through the implementation of `re-X'
strategies (e.g., refuse, rethink, design for---and implementation of---repair,
remanufacturing and recycling)~\citep{reike2018rex, eu2022ecodesign,
	eu2022repair, eu2015reman}. In addition to circular economy goals, contemporary
geopolitical tensions in an ever more globalised world have highlighted the
vulnerability of many advanced economies to intentional supply
disruptions---wrought as an act of competition or even outright
hostility~\citep{jrc2023supplychain,hartley2024cepolitics,berry2023crm}.

@@

Several LCIA methods include, to some extent, waste generation (Swiss
Eco-Factors, EDIP and
EN15804)~\citep{foen2021ecofactors,hauschild2003edip,cen2019en15804} and
material consumption (Crustal Scarcity Indicator (CSI) and Swiss
Eco-Factors~\citep{arvidsson2020csi,foen2021ecofactors}). These methods,
however, are generally limited in their scope (especially for waste), do not
allow for flexible quantification of specific waste and material types, and
often provide results in characterised units that are abstract or difficult to
interpret (e.g., \textit{Ümweltbelastungspunkte} (UBP) [English: environmental impact points] in the case of the SwissEco-Factors)~\citep{su2020sustainableproddev}.

@@

@@

In the default configuration of T-reX, there are 10 waste categories which are
then further divided by their unit of measurement (kilograms and cubic meters)
to create a total of 20 waste methods. The waste categories include
aggregations such as `total waste' `hazardous waste' as well as, those for`'
waste sent to treatment by, for example, incineration or recycling. The full
details are listed in the supplementary material in~\autoref{sec:supplementary}
[section 3]. One advantage of T-reX is that it can identify `hidden' waste
exchanges that would otherwise be `consumed' by a treatment process and not
leave the technosphere.

@@

An overview of the T-reX workflow is presented in
\autoref{fig:methods-flowchart}. Section four of the supplementary material in
\autoref{sec:supplementary} contains a more detailed computational flowchart.

\begin{figure*}
	\centering
	\caption{The workflow of T-reX presented in a generalised (a) and computational (b) format. Application of T-reX to one or more LCA databases creates a \texttt{brightway} project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}
	\includegraphics[width=1.7\columnwidth]{figures/T-reX_method.pdf}\label{fig:methods-flowchart}
\end{figure*}

@@

To provide an example, we investigated five types of Li-ion batteries in the
\texttt{ecoinvent} 3.9.1 `cutoff' LCA database~\citep{ecoinvent2016version3},
each represented by their unamended global market inventories. The system
boundaries. These following markets were used for the example calculations and
unaltered by the authors:

\begin{itemize}
	\item Li-ion, LFP, rechargeable, prismatic
	\item Li-ion, LiMn\(_2\)O\(_4\), rechargeable, prismatic
	\item Li-ion, NMC111, rechargeable, prismatic
	\item Li-ion, NMC811, rechargeable, prismatic
	\item Li-ion, NCA, rechargeable, prismatic
\end{itemize}

@@

\begin{itemize}
	\item ReCiPe 2016 v1.03, midpoint (I)~\citep{huijbregts2016recipe}
	\item Environmental Footprint (EF) v3.1 (no long-term emissions)~\citep{eu2023ef}
	\item Spatial differentiation in life cycle impact assessment---the EDIP 2003 no long
	      term emissions method~\citep{hauschild2003edip}
	\item The crustal scarcity indicator method---CSI 2020~\citep{arvidsson2020csi}
\end{itemize}

@@

An example of the output from the application of T-reX has been included in the
supplementary material in section 6 of the ~\autoref{sec:supplementary}. The
manipulated \texttt{ecoinvent} databases (which are the main product of T-reX)
can be recreated by following the instructions in the package
documentation~\citep{mcdowall2023T-reXdocs}.

@@

\autoref{fig:sankey_waste} presents a heavily abridged Sankey diagram depicting the supply chain flows of `total solid waste' attributed to the NMC 811 Li-ion battery at the global market in the \texttt{ecoinvent} database version 3.9.1. As the supply chains are expansive, this visualisation has been greatly simplified to be legible for print, multiple versions are included in the supplementary material~\autoref{sec:supplementary}. As is evident in the figure, the largest contributors to the waste footprint are the extraction and processing of the raw materials, with the largest flows coming from the extraction of cobalt, lithium, and nickel. Due to the nature of LCA, the waste flows do not represent the `real' physical flows directly, but rather an abstraction thereof based on, in this case, economic allocation---the market price of the functional unit produced by the activity~\citep{guinee2004economicallocation}. If an unallocated database were available, the waste flows could be calculated with T-reX in the same way and would be more directly related to the physical flows of the activities.

\begin{figure}[H!]
	\centering
	\includegraphics[width=0.8\columnwidth]{figures/T-reX_NMC811_WasteTotalSolid.pdf}
	\caption{A greatly simplified Sankey diagram from \texttt{ActivityBrowser showing the total solid waste footprint flows for the
	 NMC 811 battery supply chain at the global market in the \texttt{ecoinvent} database version 3.9.1}label{fig:sankey_waste}
\end{figure}

\subsubsection{Temporal and scenario variation in waste and material inventory footprints}

%\label{sec:results-case_study-total_footprints}

@@

\begin{figure*}[H]
	\centering
	\includegraphics[width=/textwidth]{figures/T-reX-wastefootprint-totalANDcarbon.pdf}
	\caption{T-reX calculated supply chain inventories for the five Li-ion batteries in the case study, showing their `footprints' for: (a) total solid waste, and (b) carbon dioxide waste (produced from carbon capture and storage (CCS) only). The footprints were modelled with prospective LCA databases from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:waste_totalANDcarbon}
\end{figure*}

@@

\begin{figure*}
	\centering
	\includegraphics[width=1.9\columnwidth]{figures/T-reX-coalANDedip.pdf}
	\caption{Comparison of the (a) `T-reX – Coal (black)' demand footprint with the (b) conventional LCIA method `EDIP 2003 – coal no longterm'. In the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model for all five batteries.}\label{fig:comparison_methods}
\end{figure*}

@@

In the most similar published method,~\cite{laurenti2023wastefootprint} used an
alternative procedure for calculating the waste footprint of \~1400 activities
in the \texttt{ecoinvent} database version 3.5. Direct comparison is not
possible, as this database contains only one generic Li-ion battery, `market
for battery, Li-ion, rechargeable, prismatic'. The inventory of this battery
most closely resembles that of the NMC 111 battery in this case study.
\autoref{tab:results-case_study-comparison} presents a comparison of the
results from the two studies, where possible. For liquid, solid, and recycled
waste, the results were closely aligned, however, for the hazardous waste
fraction, Laurenti et al. reported 95\%, whereas T-reX reported only 3\%. The
reason for this discrepancy is explained by the fact that in the method of
Laurenti et al.\, a ``waste flow was regarded as hazardous when the
hazardousness was clearly stated in its subsequent waste treatment activity''.
The authors continue: ``It should be noted that the high hazardousness ratios
for many products might indicate a weakness in the validity of this measure''.
In T-reX, the source database is deconstructed into a list of separate
exchanges, and only those explicitly defined as hazardous are marked as such.

@@

% Back matter
%\appendix
\section{Supplementary material}\label{sec:supplementary}
The supplementary material supplied with this manuscript contains the
following sections:
\begin{enumerate}
	\item Metadata of the T-reX Python package
	\item Details of the modules and user configuration
	\item Description of the computational workflow
	\item Modular flowchart of T-reX
	\item Example of the terminal output of T-reX
	\item Example of the terminal output of the case study
	\item Complete tabulated results of the case study
	\item Complete visualisations from the case study
\end{enumerate}

@@

\includegraphics[width=0.8\columnwidth]{figures/T-reX_credit_heatmap.pdf}

@@

% insert supplementary information
%\clearpage
%#\includepdf{T-reX_supplementary-material/00_T-reX_supplementary-material_main.pdf}
